---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(here)

vf <- read_csv(here::here("data/processed/voterfile_combined.csv"))
train <- read_csv(here::here("data/processed/training_data.csv"))
# only reading in responses to determine overall response rates
# only training data will be used when comparing to other variables
resp <- read_csv(here::here("data/processed/all_responses.csv"))

```

Breakdown of responses, how many with no response

```{r}
left_join(vf, resp, by = c("state", "test_id")) %>% 
  count(recall_vote, sort = TRUE)
```

Yes outnumber the nos, but we have a high rate of don't know: `r nrow(resp %>% filter(recall_vote == "DK")) / nrow(resp)`

how many people do we drop from the responses

```{r}
anti_join(resp, vf, by = c("state", "test_id")) %>% 
  count(state, sort = TRUE) %>% 
  mutate(sum(n))
```

298 are dropped, 38 because they are out of state

## breakdown of demographics in the district

missingness in the voterfile

```{r}

miss <- lapply(vf, function(x) sum(is.na(x)))

miss[miss > 0]

```

11 missing age, 9 missing household age, quite a few missing modeled household income, NA for hh_children does not mean 0 here, density cluster also missing quite a bit; could this be filled in by census tract? 

party affiliation

```{r}

count(vf, ca_partyaffiliation, sort = TRUE)
```

```{r}

count(vf, district, ca_partyaffiliation) %>% 
  spread(district, n)

```

03 heavily dem (`r 43391 / (43991 + 21990)`) while 11 more split (`r 27584 / (27584 + 20834)`)

## comparing demos to outcome

```{r}
train %>% 
  group_by(ca_partyaffiliation) %>% 
  count(recall_vote, sort = TRUE) %>% 
  mutate(pct_party = n / sum(n))

```

Reps more strongly favor recall than Dems oppose; NPA also more strongly support recall; key question is what to do with the don't knows. 

responses by district

```{r}
train %>% 
  group_by(district) %>% 
  count(recall_vote, sort = TRUE) %>% 
  mutate(dist_pct = n / sum(n))
```

Despite district 3 having higher dem pct, 1 point in favor of recall with less uncertainty

