---
title: "R Notebook"
output: html_notebook
---

```{r}
library(here)
library(corrplot)
library(ggbiplot)
library(broom)
library(tidyverse)
library(GGally)

vf <- read_csv(here::here("data/processed/voterfile_combined.csv"))
train <- read_csv(here::here("data/processed/training_data.csv"))
# only reading in responses to determine overall response rates
# only training data will be used when comparing to other variables
resp <- read_csv(here::here("data/processed/all_responses.csv"))
```

Breakdown of responses, how many with no response

```{r}
left_join(vf, resp, by = c("state", "test_id")) %>% 
  dplyr::count(recall_vote, sort = TRUE)
```

Yes outnumber the nos, but we have a high rate of don't know: `r nrow(resp %>% filter(recall_vote == "DK")) / nrow(resp)`

how many people do we drop from the responses

```{r}
anti_join(resp, vf, by = c("state", "test_id")) %>% 
  dplyr::count(state, sort = TRUE) %>% 
  mutate(sum(n))
```

298 are dropped, 38 because they are out of state

## breakdown of demographics in the district

missingness in the voterfile

```{r}

miss <- lapply(vf, function(x) sum(is.na(x)))

miss[miss > 0]

```

11 missing age, 9 missing household age, quite a few missing modeled household income, NA for hh_children does not mean 0 here, density cluster also missing quite a bit; could this be filled in by census tract? 

party affiliation

```{r}

dplyr::count(vf, ca_partyaffiliation, sort = TRUE)
```

```{r}

dplyr::count(vf, district, ca_partyaffiliation) %>% 
  spread(district, n)

```

03 heavily dem (`r 43391 / (43991 + 21990)`) while 11 more split (`r 27584 / (27584 + 20834)`)

political activity of population

```{r}

qplot(vf$es_genrate, binwidth = .05)
```

```{r}
dplyr::count(vf, es_anyprimary)
```

```{r}
qplot(vf$sy_gunscore, binwidth = .01)
```

```{r}
qplot(vf$sy_voteprop2012, binwidth = 1)
```

### ACS

```{r, message = FALSE, warning = FALSE}
# 
# train %>% 
#   select(53:68) %>% 
#   ggpairs()
```

## household



## comparing demos to outcome

```{r}
train %>% 
  group_by(ca_partyaffiliation) %>% 
  dplyr::count(recall_vote, sort = TRUE) %>% 
  mutate(pct_party = n / sum(n))

```

Reps more strongly favor recall than Dems oppose; NPA also more strongly support recall; key question is what to do with the don't knows. 

responses by district

```{r}
train %>% 
  group_by(district) %>% 
  dplyr::count(recall_vote, sort = TRUE) %>% 
  mutate(dist_pct = n / sum(n))
```

Despite district 3 having higher dem pct, 1 point in favor of recall with less uncertainty

Plotting some of the relationships

```{r}
num_plots <- train %>% 
  select_if(is.numeric) %>% 
  select(-starts_with("ac"), -1, -3, -5, -6, -recall_binary,
         -di_statehousedistrict) %>% 
  mutate(log_density = log(pop_density))

correlations <- cor(num_plots, use = "pairwise.complete.obs")

corrplot(correlations)

```

it looks like we can caputre more of the variable with three sections:

- household info
- acs info
- ideology

some of the modeled variables are correlated with acs info

In addition, recall most strongly correlated with partisan vote score and ideaology, so probably ony need one of those

state house district has a strong relationship with race

## PCA

```{r}

pca_comp <- num_plots %>% 
  filter(!is.na(hh_children), !is.na(bach_or_higher), !is.na(log_density))

dat_pca <- prcomp(pca_comp, center = TRUE, scale. = TRUE)

summary(dat_pca)
```


```{r}

tidied_pca <- broom::tidy(dat_pca$rotation) %>% 
  gather(pc, contribution, PC1:PC36)

tidied_pca %>% 
  filter(pc == "PC1") %>% 
  ggplot(aes(reorder(.rownames, contribution), contribution)) +
    geom_col() +
    coord_flip()
```

again, it looks like houshold info and acs data on education are contributing to the most amount of variation

```{r}
tidied_pca %>% 
  filter(pc == "PC2") %>% 
  ggplot(aes(reorder(.rownames, contribution), contribution)) +
    geom_col() +
    coord_flip()
```

race starts to play a higher role here

```{r}
tidied_pca %>% 
  filter(pc == "PC3") %>% 
  ggplot(aes(reorder(.rownames, contribution), contribution)) +
    geom_col() +
    coord_flip()
```


